The role of aneuploidy in the evolution of cancer drug resistance

### 2024-02-19 EvolutionaryRescue1.tex
edits and comments

\paragraph{Evolutionary model.}
We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below. 
We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
To model standing genetic variation, we assume that before the onset of therapy, wildtype cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment (we assume that aneuploidy significantly increases the death rate of wildtype cancer cells in an environment which is not affected by chemotherapeutic drugs). % TODO unclear how s affects \lambda_w; I do not think that s has any effect on \lambda_w as s is the fitness of the aneuploid in the drug free environment
See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes.

@@

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
The first line describes the case in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wild type. 
In the second case, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.  % TODO that's a weird sentence; a lineage starts with a single cell, if the lineage size stays constant that it stays one. so what are we talking about? try to rephrase. ok
In both of these cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
In the third case, aneuploid cells are sufficiently resistance for the cancer population to re-grow the tumor even without additional resistance mutations.
Notably, in this case the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
In all cases, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second case where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A).  % TODO not sure what the last sentence "but ..." means and how do you see it ok
% TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4

@@

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, pushing the system to the second case. In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation. % NOTE this is the kind of conclusions we want to drive home
This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.

@@

In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
However, aneuploidy is likely produced even before onset of treatment at some rate $\tilde{u} \le u$ and confer a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}, and the aneuploidy rate in the presence of drugs, $u$, is likely higher than in their absence, $\tilde{u}$ \citep{wang2019molecular,mason2017functional}.
But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ (we assume that the division rate of the wildtype in a drug-free environment is equal to the division rate of the wildtype in the environment affected by the anti-cancer drug) of aneuploid cells in the population.

Therefore, the ratio of threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that 
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}
Comparing the above approximation to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).  % TODO which approximations? unclear ok
Standing genetic variation will drive evolutionary rescue to the drug if wildtype growth rate $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if the aneuploidy cost $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost, $s$, is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\times10^{-3}$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 1.825$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.

Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for direct mutation,
\begin{equation} \label{eq:N_sgv_ratio}
\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
    \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{\tilde{u}}\left(v \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
We observe that evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $s$ is very large or the effect of the drug $\Delta_w$ is small.  In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegreagation rate $\tilde{u}$ is very large. The ratio does not depend on the rate of chromosome missegregation induced by the drug $u$. However, if the aneuploid growth rate $\Delta_a$ increases, then evolutionary rescue is driven by standing genetic variation. For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.356$, which means that standing genetic variation reduces the threshold tumor size by approximately $65\%$. We observe that while standing genetic variation does not drive evolution of drug resistance when compared to de-novo aneuploidy, it does offer an advantage when compared with direct mutation.

@@

When evolutionary rescue occurs, the time until recurrence of the tumor may still be very long. We therefore explored the time until recurrence of the tumor, that is, the time until the tumor reaches its original size, $N$.
When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an even occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
However, when the expected number of resistant lineages is large, the behavior of the size of the population of mutant cells is deterministic (i.e. it is modeled by a system of ODEs, see \Cref{detODE}) and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size. % TODO I dont like the word "behaves", please be specific. - what exactly is deterministic and why does it mean you cannot do the separation. ok
Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .

\paragraph{Evolutionary rescue time.}
We have derived approximations for $\tau_m$, the mean evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the mean rescue time with aneuploidy ($u>0$), both conditioned on evolutionary rescue occurring (see appendix \ref{sec:appendix_rescue_time}).
These approximations are in good agreement with simulation results for small, intermediate, and large tumor sizes (\Cref{EvolutionaryRescueTimeComplete,MeanTimeGrowthAneuploidyPlot}).
The mean rescue time with aneuploidy for small and large tumors follows these expressions:
\begin{equation}  \label{eq:AsymptoticTimeRules}
\tau_a \approx \begin{cases}
    \frac{1}{\abs{\Delta_w}}+\frac{1}{\abs{\Delta_a}} ,&
 N \ll N_a^* ,\\ 
  \frac{1}{v\lambda_w N}   \frac{\lambda_m}{\Delta_m} ,& % I replaced pm with 𝜆𝑚 − 𝜇𝑚 / 𝜆𝑚
  N \gg N_m^* .
  \end{cases}
\end{equation}

@@

When a fraction $f\approx 1.36\%$ of the initial cancer cell population is aneuploid we calculate the mean evolutionary rescue time in \cref{meantimet2SGV} which we plot in \Cref{SGVEvolutionaryRescueTimeComplete}. We note that standing genetic variation reduces the mean evolutionary rescue time only for intermediary sized tumors but not for large or small tumors.
\paragraph{Recurrence time.}

In appendix \ref{sec:appendix_recurrence_time} we approximate the mean time for the population of mutant cancer cells to reach the initial population size $N$ which we denote as the recurrence time. \Cref{proliferationFigure} shows the agreement between our approximations and simulations. For values of $N$ smaller then $10^8$ the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$. For $N>10^8$ the behavior of the mutant cancer cell population is deterministic and the mean recurrence time becomes constant. We note that $\tau_a^r$ has the following asymptotic behavior:
\begin{equation} \label{eq:AsymptoticTimeRulesRecurrence}
\tau_a^r \approx \begin{cases}
   \frac{1}{ \abs{\Delta_w}}+\frac{1}{ \abs{\Delta_a}}+\frac{\log p_mN}{\Delta_m} ,&
 N \ll N_a^* ,\\ 
  \frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}  ,&
  N \gg N_m^* .
  \end{cases}
\end{equation}
In the first case, when tumor is small, the mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time (\Cref{ProliferationTimeLarge}). Increasing the mutant growth rate $\Delta_m$ leads to shorter recurrence times while increasing the wildtype and aneuploid growth rates $\Delta_w$ and $\Delta_a$, retrospectively, increases the recurrence time. In the second case, when tumor is large, we observe that increasing either the mutant growth rare $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time needed for the tumor to rebound to its initial size. Additionally, drugs which significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delays cancer relapse. Consequently, patients which take such drugs require a longer period of monitoring to guarantee the effectiveness of the treatment.
\paragraph{Distribution of evolutionary rescue time with and without aneuploidy.}

In Appendix E we derive the probability that by time $t$ a successful mutant has been generated. \Cref{cdffig}A show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent.  We observe that aneuploidy start to have an effect on adaptation for timescales greater then 100 days.

\paragraph{Distribution of  recurrence time.}
In Appendix F we derive the probability that mutant cancer cell population reaches $N$ at time $t$. \Cref{cdffig}B shows agreement between our approximations and stochastic simulations for various values of $N$. Additionally, we derived the distribution of the recurrence time and plotted our result in \Cref{KMdistribution} alongside simulations for the case $N=10^6$ (small tumor) and we note that the distribution is wide and rights-skewed. It is highly unlikely to observe the recurrence of tumors at times smaller then $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}\approx1542$ days for the parameter values $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ and independent of initial tumor size $N$ (\Cref{cdffig}B).

The recurrence can also be defined as the time necessary for the tumor size to reach a detection threshold $M$. We derive the mean recurrence time to detection size $M=10^7$ in appendix \ref{sec:appendix_recurrence_time} and plot in \Cref{RecurrencePlot} with simulations. We observe that  for small and intermediate sized tumors the effect of the detection size is negligible when compared to $\tau_a^r$. However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$.

Most clinical trials report data on the distribution of recurrence time measured from a varying time of surgery~\citep{avanzini2019cancer} with chemotherapy usually following after. Therefore, since only undetected secondary tumors are present at the time of the administration of the anti-cancer drug we lack knowledge of the size of the tumors and we cannot compare the empirical distributions to our predictions. However, we expect that the variability of secondary tumors to average out across large cohorts of patients and, as a result, we can use the mean recurrence time. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

>>>>

\paragraph{Evolutionary model.}
We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below. 
We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
To model standing genetic variation, we assume that before the onset of therapy, wildtype cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to wildtype cells in a drug-free environment.
See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes.

@@

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
The first line describes the case in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wild type. 
In the second case, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
In both of these cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
In the third case, aneuploid cells are sufficiently resistance for the cancer population to re-grow the tumor even without additional resistance mutations.
Notably, in this case the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
In all cases, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second case where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A). % TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4

@@

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, pushing the system to the second case. In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation.
This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.

@@

In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
However, aneuploidy is likely produced even before onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}. Furthermore, the aneuploidy rate in the presence of drugs is likely higher than in their absence, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but no the division rate and therefre use $\lambda_w$ for the wildtype division rate in the drug-free enviornment).

Therefore, the ratio of threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost $s$, such that 
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}
Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
Standing genetic variation will drive evolutionary rescue to the drug if wildtype growth rate $\Delta_w$ is very negative due to a strong effect of the drug on sensitive cells, or if the cost aneuploidy cost in the drug-free environment, $s$, is small.
In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\times10^{-3}$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 1.825$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.

Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for direct mutation,
\begin{equation} \label{eq:N_sgv_ratio}
\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
    \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{\tilde{u}}\left(v \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $s$ is very large or the effect of the drug $\Delta_w$ is small.  In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegreagation rate $\tilde{u}$ is very large. The ratio does not depend on the rate of chromosome missegregation induced by the drug $u$. However, if the aneuploid growth rate $\Delta_a$ increases, then evolutionary rescue is driven by standing genetic variation. For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.356$, which means that standing genetic variation reduces the threshold tumor size by approximately $65\%$. Therefore, standing genetic variation does not drive evolution of drug resistance when compared to de-novo aneuploidy, but it does offer an advantage when compared with direct mutation.

@@

When evolutionary rescue occurs, the time until recurrence of the tumor may still be very long. We therefore explored the time until recurrence of the tumor, that is, the time until the tumor reaches its original size, $N$.
When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an even occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
However, when the expected number of resistant lineages is large, the dynamics of number of mutant cells is deterministic (i.e., it can be modeled by a system of ODEs, \cref{detODE}) and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size. % TODO I dont like the word "behaves", please be specific. - what exactly is deterministic and why does it mean you cannot do the separation. ok
Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .

\paragraph{Evolutionary rescue time.}
We have derived approximations for $\tau_m$, the mean evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the mean rescue time with aneuploidy ($u>0$), both conditioned on evolutionary rescue occurring.
These approximations are in good agreement with simulation results for small, intermediate, and large tumor sizes (\Cref{EvolutionaryRescueTimeComplete,MeanTimeGrowthAneuploidyPlot}).
The mean rescue time with aneuploidy for small and large tumors follows these expressions (Appendix \ref{sec:appendix_rescue_time}),
\begin{equation}  \label{eq:AsymptoticTimeRules}
\tau_a \approx \begin{cases}
    \frac{1}{\abs{\Delta_w}}+\frac{1}{\abs{\Delta_a}} ,&
 N \ll N_a^* ,\\ 
  \frac{1}{v\lambda_w N}   \frac{\lambda_m}{\Delta_m} ,& % I replaced pm with 𝜆𝑚 − 𝜇𝑚 / 𝜆𝑚
  N \gg N_m^* .
  \end{cases}
\end{equation}

@@

When a fraction $f\approx 1.36\%$ of the initial cancer cell population is aneuploid, the mean evolutionary rescue time (\cref{meantimet2SGV}; \Cref{SGVEvolutionaryRescueTimeComplete}. We note that standing genetic variation reduces the mean evolutionary rescue time only for intermediary sized tumors but not for large or small tumors. % WHAT is the purpose of this sentence?

In Appendix E we derive the probability that by time $t$ a successful mutant has been generated. \Cref{cdffig}A show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent.  We observe that aneuploidy start to have an effect on adaptation for timescales greater then 100 days. % WHAT is the purpose of this sentence?

\paragraph{Recurrence time.}

We next approximate the mean time for the population of mutant cancer cells to reach the initial, pre-treatment population size $N$, which we denote the recurrence time $\tau_a^r$ (Appendix \ref{sec:appendix_recurrence_time}),
\begin{equation} \label{eq:AsymptoticTimeRulesRecurrence}
\tau_a^r \approx \begin{cases}
   \frac{1}{ \abs{\Delta_w}}+\frac{1}{ \abs{\Delta_a}}+\frac{\log p_mN}{\Delta_m} ,&
 N \ll N_a^* ,\\ 
  \frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}  ,&
  N \gg N_m^* .
  \end{cases}
\end{equation}
\Cref{proliferationFigure} shows the agreement between our approximations and simulations.
For small tumors ($Nt \ll 10^8$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$. The mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time. Increasing the mutant growth rate $\Delta_m$ leads to shorter recurrence times while increasing the wildtype and aneuploid growth rates, $\Delta_w$ and $\Delta_a$ respectively, increases the recurrence time.
For large tumors ($N \gg 10^8$), the dynamics of the number of mutant cells is deterministic and the mean recurrence time becomes constant.
Increasing either the mutant growth rate $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time for the tumor to rebound to its initial size.
In addition, drugs that significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delay cancer relapse.
Consequently, patients treated with such drugs require a longer period of monitoring to guarantee the effectiveness of the treatment.
% Remus the next paragraphs look like you wanted to find approximations so you did, you reference them here, but the results are buried in the technical text: WHAT IS IT THAT YOU WANT US TO LEARN from this? 
In Appendix F we derive the probability that mutant cancer cell population reaches $N$ at time $t$. \Cref{cdffig}B shows agreement between our approximations and stochastic simulations for various values of $N$. Additionally, we derived the distribution of the recurrence time and plotted our result in \Cref{KMdistribution} alongside simulations for the case $N=10^6$ (small tumor) and we note that the distribution is wide and rights-skewed. It is highly unlikely to observe the recurrence of tumors at times smaller then $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}\approx1542$ days for the parameter values $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ and independent of initial tumor size $N$ (\Cref{cdffig}B).

The recurrence can also be defined as the time necessary for the tumor size to reach a detection threshold $M$. We derive the mean recurrence time to detection size $M=10^7$ in appendix \ref{sec:appendix_recurrence_time} and plot in \Cref{RecurrencePlot} with simulations. We observe that  for small and intermediate sized tumors the effect of the detection size is negligible when compared to $\tau_a^r$. % WHAT DOES THAT MEAN?
However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$.

Most clinical trials report data on the distribution of recurrence time measured from a varying time of surgery~\citep{avanzini2019cancer} with chemotherapy usually following after. % WHAT DOES "varying" mean here?
Therefore, since only undetected secondary tumors are present at the time of the administration of the anti-cancer drug we lack knowledge of the size of the tumors and we cannot compare the empirical distributions to our predictions. However, we expect that the variability of secondary tumors to average out across large cohorts of patients and, as a result, we can use the mean recurrence time. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
